---
title: "Class 5: Data Viz with ggplot"
author: "Rachel Galleta (PID:A16859649)"
format: gfm
---
## introduction

Today we are exploring the **gggplot** package and how to make nice figures in R.

There are lots of ways to make figures and plot in R. These include:

- so called "base" R
- and add on packages like **ggplot2**

Here is a simple "base" R plot.

```{r}
head(cars)
```

We can simply pass to to the `plot()` function.

```{r}
plot(cars)
```

> Key-point: Base R is quick but not so nice looking in some folks eyes.

Let's see how we can plot this with **ggplot2**...

1st I need to install this add-on package. For this we use the `install.packages()` function - **WE DO THIS IN THE CONSOLE, NOT our report**. This is a one time only deal.

2nd We need to load the package with the `library()` function every time we want to use it. 


```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot is composed of at least 3 layers:

- **data** (i.e a data.frame with the things you want to plot),
- aesthitics **aes()** that map the columns of data to your plot features (i.e. aesthitics)
- geoms like **geom_point()** that srt how the plot appears

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()
```

> Key point: For simple "canned" graphs base R is quicker but as things get more custom and elobrate then ggplot wins out...

Let's add more layers to our ggplot

Add a line showing the relationship between x and y
Add a title
Add custom axis labels "Speed (MPH)" and "Distance (ft)"
Change the theme...

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  labs(title="Silly plot of Speed vs Stoping distance",
       x="Speed (MPH)",
       y="Distance (ft)") +
  theme_bw()
```

## Going further

Read some gene expresion data

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)

head(genes)
```

> Q1. How many genes are in this wee dataset?

```{r}
nrow(genes)
```
```{r}
ncol(genes)
```

> Q2. How many "up" regulated genes are there?

```{r}
sum( genes$State == "up" )
```

A useful function for counting up occurances of things in a vector is the `table()` function.


```{r}
table( genes$State )
```

Make a v1 figure

```{r}
p <- ggplot(genes) +
  aes(x=Condition1,
      y=Condition2,
      col=State) +
  geom_point() 

p
```

```{r}
p + 
  scale_colour_manual( values=c("blue","gray","red") ) +
  labs(title="Expresion changes upon drug treatment",
       x="Control (no drug)",
       y="Treatment (with drug)") +
  theme_bw()
```

## More Plotting 

Read in the gapminder dataset

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

Lets have a wee peak

```{r}
head( gapminder, 3)
```

```{r}
tail(gapminder, 3)
```


> Q4. How many different country values are in this dataset?

```{r}
nrow(gapminder)
```
```{r}
length( table(gapminder$country) )
```
> Q5. How many different continent values are in this dataset.

```{r}
unique(gapminder$continent)
```
```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point()
```

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent, label=country) +
  geom_point() +
  geom_text()
```

I can use the **ggrepl** package to make more sensible labels here.

```{r}
library(ggrepel)

ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent, label=country) +
  geom_point() +
  geom_text_repel()
```

I want a seperate pannel per continent

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent, label=country) +
  geom_point() +
  facet_wrap(~continent)
```

## Summary

The main advantages of ggplot over base R plot are:

1. **Layered Grammar of Graphics**: ggplot uses a consistent, layered approach where you build plots by adding layers for data, aesthetics, and geometries. This makes complex plots easier to construct and modify, compared to base R which often requires different functions and many arguments for each plot type [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

2. **Publication-Quality Output**: ggplot produces visually appealing, publication-ready figures by default, while base R plots often need extensive tweaking to look polished [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

3. **Declarative Syntax**: You specify what you want to show (data, mappings, geoms) rather than how to draw each element, making code more readable and maintainable [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

4. **Easy Customization and Extension**: Adding titles, axis labels, legends, colors, and themes is straightforward and consistent. In base R, these customizations can be fiddly and require more manual control [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

5. **Reproducibility**: ggplot code is easy to reuse and adapt for new datasets or visualizations, supporting reproducible research [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

Which of these advantages do you think is most important for your work? Why?

