---
title: "class 6"
author: "Rachel Galleta(A16859649)"
format: gfm
---
All functions in R have at least 3 things:
-A **name**, we pick this and use it to call the function.

-input **arguments**, there can be multiple comma separated inputs to the function.
-the **body**, lines of R code that do the work of the function.

our first function:

```{r}
add<- function(x, y=1) {  
  x+y
  }
```
lets test ot function 

```{r}
add(c(1,2,3),y=10)
```


```{r}
add(10)
```
```{r}
add(10,10)
```
## A second function

Lets try something more interesting.Make a sequence generation tool.

the 'sample()' fucntion coulb be useful here.
```{r}
sample(1:10,size=3)
```
change this to work with nucleotides A,C,G and T return 3 of them.

```{r}
n<- c("A","C","G","T")
sample(n, size=15, replace= TRUE)
```
turn this snipet into a fuction that returns a user specified legnth dna seqeunce, lets call it 'generate_dna()'

```{r}
generate_dna <-function(len=10, fasta=FALSE) {
 n<- c("A","C","G","T")
 v<-sample(n, size=len, replace= TRUE)
 
 # make a single element vector
 s<-paste(v, collapse ="")
 
 cat("well done you!")
 if(fasta){
   return(s)
 }else {
   return (v)
 }
}
```



```{r}
generate_dna(5)
```

```{r}
s<- generate_dna(15)
s
```

I want the option to return a single element charater vector with my seqence all together like this;"GGAGTAG"

```{r}
s
paste(s,collapse = "")
```
```{r}
generate_dna(10,fasta=FALSE)
```

## A more advanced example

make a thrid function that generates protein sequence of a user specified length and format.


```{r}
generate_protein<-function(size =15,  fasta=TRUE){
  aa <- c("A","R","N","D","C","Q","E","G",
          "H","I","L","K","M","F","P","S",
          "T","W","Y","V")
  seq<-sample(aa,size = size,replace=TRUE)
  
  if(fasta){
    return(paste(seq, collapse=""))
  }else{
    return(seq)
  }
  }
```

try this out...
```{r}
generate_protein(10,fasta=T)
```

>Q. generate random protein seqeunces between lenghts 5 and 12 amino-acids. 

```{r}
generate_protein(5)
generate_protein(6)
```

one approach is to do this by brute force calling our function for each length 5 to 12.

another approach is to write a 'for c' loop to litetire  the input valued 5 to 12.

a very useful third R specific approach is to use the 'supply()' function

```{r}
seq_length<-6:12
for (i in seq_length){
  cat(">",i, "\n",sep="")
  cat(generate_protein(i))
  cat("\n")
}
```

```{r}
sapply(5:12,generate_protein)
```
> **key point writting funtion with r is doable, but not the easiest thing. Starting with a working snippet of code and then using LLM tool to improve and generalize your function codes is a productive approach.


